---
title: "625 Project model fit"
output: pdf_document
date: "2023-11-02"
---

```{r}
library(haven)
library(tidyverse)
library(dplyr)
birthweight=read_xpt("P_ECQ.XPT")
demo=read_xpt("P_DEMO.XPT")
activity=read_xpt("P_PAQY.XPT")
diet=read_xpt("P_DBQ.XPT")
foodsec=read_xpt("P_FSQ.XPT")
body=read_xpt("P_BMX.XPT")

```
In this proposed model, we assume a linear relationship between the outcome and predictors. This assumption can be checked by the residual vs fitted plot. Second, linear regression requires multivariate normality among the variables, and this assumption can be checked by the normal Q-Q plot. Another assumption is that residual errors have a distribution centered around 0 with constant variance and are independent with each other and to the predictors. This is checked by the residual plot and scale location plot.

Model 1:\
$GrowthPercentile = \beta_0 + \beta_1Sex + \beta_2Race + \beta_3BirthWeight + \beta_4Smoke + \beta_5FoodSecurity + \beta_6FastFood$\
$+ \beta_7Activity + \epsilon$

Model 2:\
$sqrt(GrowthPercentile) = \beta_0 + \beta_1Sex + \beta_2Race + \beta_3BirthWeight + \beta_4Smoke + \beta_5FoodSecurity + \beta_6FastFood + \beta_7Activity + \beta_8Sex*FoodSecurity + \epsilon$

each persons bmi under conditions, each person can have different
age on bmi is different for individual
random slope before and after age 60

one random slope: inter individual variation is same before and after
inter individual differences different 

```{r}
data <- readRDS('clean_data.RDS')
data
colnames(data)
data$race <- factor(data$race, levels = c(3, 1,2,4,5,6,7))
data$activity1 <- as.character(data$activity)

# Categorize activity into levels
data <- data %>%
  mutate(activity1 = case_when(
    activity1 =='0' ~ "0",
    activity1 =='1' ~ "1-3",
    activity1 =='2' ~ "1-3",
    activity1 =='3' ~ "1-3",
    activity1 =='4' ~ "4-6",
    activity1 =='5' ~ "4-6",
    activity1 =='6' ~ "4-6",
    activity1 =='7' ~ "7",
  ))

data$activity1 <- as.factor(data$activity1)
data
```

```{r}
#Fit a linear regressiion model
m1 <- lm(sqrt(growth_percentile) ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1, data=data)
summary(m1)
AIC(m1)
plot(m1$residuals)
```

```{r}
#check for potential interactions
ggplot(data, aes(x=race, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
ggplot(data, aes(x=birthweight, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
ggplot(data, aes(x=smoke, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
ggplot(data, aes(x=food_sec, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
ggplot(data, aes(x=fastfood, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
ggplot(data, aes(x=activity1, y=growth_percentile, fill=sex)) + 
    geom_boxplot()
```

```{r}
#linear regression model with interaction
m2 <- lm(growth_percentile ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1+sex*food_sec, data=data)
summary(m2)
AIC(m2)
plot(m2)

#Added in transformation for normal distributed residuals
m3 <- lm(sqrt(growth_percentile) ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1+sex*food_sec, data=data)
summary(m3)
AIC(m3)
plot(m3)

#nrow(data)
#nrow(model.matrix(m1))
#nrow(model.matrix(m3))
#anova(m3,m1)
```

```{r}
plot(m2)
plot(m1)
```
```{r}
qqnorm(residuals(m2))
qqline(residuals(m2))
```

```{r}
#Logistic regression for obese binary variable
data <- data %>%
  mutate(growth_percentile1 = case_when(
    growth_percentile < 95 ~ 0,
    growth_percentile >= 95 ~ 1
  ))

data$growth_percentile1 <- as.factor(data$growth_percentile1)
data


m4 <- glm(growth_percentile1 ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1+sex*food_sec, data = data, family = binomial)
#m4 <- glm(growth_percentile1 ~ race+birthweight+smoke+activity1, data = data, family = binomial)
summary(m4)
AIC(m4)
plot(m4)

cooks_distance <- cooks.distance(m4)
plot(cooks_distance, pch = 20, main = "Cook's Distance Plot")
library(car)
crPlots(m4)
vif(m4)

deviance_residuals <- residuals(m4, type = "deviance")
plot(predict(m4, type = "response"), deviance_residuals, 
     ylab = "Deviance Residuals", xlab = "Fitted Values")

```

