---
title: "Analyzing the Factors Affecting BMI Percentiles and the Log Odds of Obesity of Children Aged 2-15"
author: "Aanya Mohan, Elena Chun, Siwei Cao"
date: "2023-12-10"
output:
  pdf_document: default
  word_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction:

Childhood obesity, also known as pediatric obesity, is a pressing issue
that affects youth throughout the country. According to the Center for
Disease Control (CDC), as of 2022, childhood obesity affects 1 in 5
children in the United States${_1}$. There are a variety of factors that
can lead to childhood obesity, from lifestyle to medication, or genetic
factors. Studies show that there may be a correlation between birth
weight and the likelihood of childhood obesity. Despite sex differences,
a higher birth weight increases the chances of developing obesity in
adolescents. The National Health and Nutrition Examination Survey
(NHANES) is "designed to address the health and nutritional status of
adults and children in the United States"${_2}$. This survey evaluates
all aspects of a person's health and nutritional habits. Factors
recorded such as gender, age, food security, days physically active,
mother smoked while pregnant, and number of meals from a fast food or
pizza place, can be used to perform a regression analysis. This analysis
may provide an explanation for childhood obesity trends, as defined by
birth weight.

## Data Overview:

Data for this analysis comes from the National Health and Nutrition
Examination Survey (NHANES) conducted by the CDC for 2017-2020. The
response variable for this analysis is BMI percentile and the predictor
of interest is weight at birth, measured in pounds, reported for
children ages 2-15 at the time of survey. Other predictor variables
include: gender, days physically active for at least 60 minutes out of
the week, whether mother smoked while pregnant, household food security,
race, and number of meals from a fast food or pizza place in a week.
These are all factors that are believed to explain trends in weight, and
its effect on childhood obesity. Gender and race are demographic
information reported by the survey participants. Days physically active
is a metric measuring how many days of the week youths aged 2-15 are
physically active for at least 60 minutes. Whether mother smoked while
pregnant is a binary variable that assesses the smoking habits of the
participants' biological mother. Food security is a categorical variable
which rates the household food security on a scale of 1 to 4, where 1 is
full food security and 4 is very low food security. Number of meals from
a fast food or pizza place is the estimated number of meals not prepared
at home, from a fast food restaurant or a pizza place, in a week.

## Data Cleaning:

Initially, we wanted to use BMI as the response variable for this
analysis as BMI is commonly used as an indicator of obesity. However
upon further research, we discovered that BMI percentile would be more
appropriate as children are usually measured in percentiles and a
healthy BMI can change depending on what age children are in
development. To create the BMI percentile variable, we used equations
provided by the CDC along with the variables BMI, age, and gender which
were provided in the dataset. The birth weight variable in the dataset
was given as two variables of birth weight in pounds (lb) and birth
weight in ounces (oz), which is typically how birth weight is measured
for babies. For easier interpretation, both variables were combined and
reported solely as birth weight in pounds (lb). Lastly, days of physical
activity was binned into categories of 0 days, 1-3 days, 4-6 days, and 7
days as we did not expect big differences in BMI percentiles for each
day of physical activity.

```{r, echo=FALSE,message=FALSE,fig.height=3,fig.width=3,fig}
library(haven)
library(tidyverse)
library(dplyr)
library(car)
library(gtsummary)
library(ggplot2)
data <- readRDS('clean_data.RDS')
data$race <- factor(data$race, levels = c(3, 1,2,4,5,6,7))
data$activity1 <- as.character(data$activity)

# Categorize activity into levels
data <- data %>%
  mutate(activity1 = case_when(
    activity1 =='0' ~ "0",
    activity1 =='1' ~ "1-3",
    activity1 =='2' ~ "1-3",
    activity1 =='3' ~ "1-3",
    activity1 =='4' ~ "4-6",
    activity1 =='5' ~ "4-6",
    activity1 =='6' ~ "4-6",
    activity1 =='7' ~ "7",
  ))

data$activity1 <- as.factor(data$activity1)

complete_data=na.omit(data)
complete_data$race=as.character(complete_data$race)
complete_data$food_sec=as.character(complete_data$food_sec)
complete_data$sex=as.character(complete_data$sex)
complete_data$smoke=as.character(complete_data$smoke)

#fixing variables so plot labels are correct 
#adding descriptions for race
complete_data$race[complete_data$race=="1"]="Mexican American"
complete_data$race[complete_data$race=="2"]="Other Hispanic"
complete_data$race[complete_data$race=="3"]="Non-Hispanic White"
complete_data$race[complete_data$race=="4"]="Non-Hispanic Black"
complete_data$race[complete_data$race=="6"]="Non-Hispanic Asian"
complete_data$race[complete_data$race=="7"]="Other Race (Multi)"

#adding descriptions for food security
complete_data$food_sec[complete_data$food_sec=="1"]="Full Security"
complete_data$food_sec[complete_data$food_sec=="2"]="Mariginal Security"
complete_data$food_sec[complete_data$food_sec=="3"]="Low Security"
complete_data$food_sec[complete_data$food_sec=="4"]="Very Low Security"

#adding descriptions for gender
complete_data$sex[complete_data$sex==1]="Male"
complete_data$sex[complete_data$sex==2]="Female"

#adding descriptions for smoke
complete_data$smoke[complete_data$smoke==1]="Yes"
complete_data$smoke[complete_data$smoke==2]="No"

#factoring all categorical variables
complete_data$race=as.factor(complete_data$race)
complete_data$food_sec=as.factor(complete_data$food_sec)
complete_data$sex=as.factor(complete_data$sex)
complete_data$smoke=as.factor(complete_data$smoke)

```

## Exploratory Analysis:

To explore potential relationships between BMI percentile and various
covariates, we conducted exploratory analysis to solidify which
variables would be included in our model. Previous literature has
suggested that there is a positive correlation between BMI percentile
and birth weight. In the initial plot of BMI percentile and birthweight
(Figure 1A), it is unclear whether there is a relationship between the
two variables due to the dense data. After taking a random sample of 50
observations from the dataset (Figure 1B), we can see a slight positive
correlation between BMI percentile and birth weight. Additionally, other
notable covariate relationships with BMI percentile include race,
household food security, and activity. In Figure 1C, differences in BMI
percentiles between race are shown and Non-Hispanic Asian children have
noticeably lower average BMI percentiles than other races. Differences
in BMI percentiles and household food security are shown in Figure 1D
and BMI percentiles appear to increase as household food security
decreases. A similar relationship was found between BMI percentile and
days of physical activity where BMI percentiles decreased as days of
physical activity increased. Children with mothers who smoked during
pregnancy had higher BMI percentiles than those who did not. Other
covariates that did not show discernible differences in BMI percentile
were gender and number of meals from fast food place. However, these
covariates were included in the analysis as we expect these variables to
influence BMI percentile based on previous literature.

### Figure 1

```{r, echo=FALSE,warning=FALSE}
library(ggpubr)
random_sub_data=sample_n(data,50)
par(mfrow = c(1, 2))
bw_plot=ggplot(data,aes(x = birthweight, y = growth_percentile)) + geom_point(alpha=0.5) + ylab("Growth Percentile") + xlab("Birth Weight (lbs)") + scale_x_continuous(breaks = seq(3,11,1)) + labs(subtitle = "Plot of Growth Percentile vs. Birth Weight(lbs)") + theme(plot.subtitle = element_text(size = 9))

random_plot=ggplot(aes(x=birthweight,y=growth_percentile),data=random_sub_data)+geom_point()+labs(subtitle = "Scatterplot of BMI Percentile vs. Birthweight") + ylab("BMI Percentile") + xlab("Birthweight (lb)") + xlim(4,10)+ theme(plot.subtitle = element_text(size = 9))

race_boxplot=ggplot(aes(x=race, y=growth_percentile,fill=race),data=complete_data) + geom_boxplot()+labs(subtitle = "Boxplot of BMI Percentile vs. Race",fill="Race") + ylab("BMI Percentile") + xlab("Race")+ theme(plot.subtitle = element_text(size = 9))

food_sec_boxplot=ggplot(aes(x=food_sec, y=growth_percentile,fill=food_sec),data=complete_data) + geom_boxplot() + labs(subtitle = "Boxplot of BMI Percentile vs. Food Security",fill="Food Security") + ylab("BMI Percentile") + xlab("Household Food Security")+ theme(plot.subtitle = element_text(size = 9))

ggarrange(bw_plot,random_plot,race_boxplot+rremove("x.text"),food_sec_boxplot+rremove("x.text"),labels = c("A","B","C","D"))

```

## Model:

The exploratory analysis of this data resulted in the following initial
linear model:\
$BMI Percentile = \beta_{0} + \beta_{1}Sex + \beta_{2}Race + \beta_{3}BirthWeight + \beta_{4}Smoke + \beta_{5}FoodSecurity + \beta_{6}FastFood + \beta_{7}Activity + \varepsilon$

After fitting this model, we were interested in whether the relationship
between BMI percentile and a predictor depends on levels of another
predictor. To find these interactions, we plotted boxplots of the
relationships between the response and a predictor against levels of
another predictor. We found that the association between BMI percentile
and food security is not constant across males and females. Thus, we
included the interaction between food security and sex into our
regression model:\
$BMI Percentile = \beta_{0} + \beta_{1}Sex + \beta_{2}Race + \beta_{3}BirthWeight + \beta_{4}Smoke + \beta_{5}FoodSecurity + \beta_{6}FastFood$\
$+ \beta_{7}Activity + \beta_{8}Sex*FoodSecurity +\varepsilon$

However, the resulting normal qq plot deviated from a linear line,
meaning the normality assumption of residuals was not fulfilled. Thus,
we added in a square root transformation to the BMI percentile response
variable. The resulting linear regression model is:\
$sqrt(BMI Percentile) = \beta_{0} + \beta_{1}Sex + \beta_{2}Race + \beta_{3}BirthWeight + \beta_{4}Smoke + \beta_{5}FoodSecurity$\
$+\beta_{6}FastFood + \beta_{7}Activity + \beta_{8}Sex*FoodSecurity +\varepsilon$

This linear regression model allows us to identify the factors
associated with BMI percentiles of children. However, we would also like
to examine factors associated with child obesity. The CDC identifies BMI
percentiles at or above the 95th percentile as obese, so we accordingly
transformed our BMI percentile response variable to a binary variables
(0: not obese, 1: obese). We fit a logistic regression model of this
obesity binary variable to the predictors we identified in our linear
regression model:\
$P(obesity = 1) = \frac{1}{1 + e^{-(\beta_0 + \beta_1Sex + \beta_2Race + \beta_3BirthWeight + \beta_4Smoke + \beta_5FoodSecurity + \beta_6FastFood + \beta_7Activity)}}$

```{r, include=FALSE}
library(pacman)
library(boot)
#Logistic regression for obese binary variable
data <- data %>%
  mutate(obese = case_when(
    growth_percentile < 95 ~ 0,
    growth_percentile >= 95 ~ 1
  ))

data$gobese <- as.factor(data$obese)
data

m4 <- glm(obese ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1, data = data, family = binomial)
summary(m4)
AIC(m4)
m3 <- lm(sqrt(growth_percentile) ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1+sex*food_sec, data=data)

```

## Diagnostics:

First, we will examine diagnostics of the linear regression model.The
residuals vs. fitted values plot has a random scatter, indicating that
the linearity assumption is held. The normal Q-Q plot follows the linear
line relatively well, showing that the residuals are approximately
normally distributed.

### Figure 2

```{r, echo=FALSE,fig.width=6, fig.height=3}
par(mfrow = c(1, 3))
plot(m3, which=c(1,2,3))
```

The scale-location plot has relatively equally spread points across the
line, suggesting that the assumption of constant variance is reasonable.
This shows that the linear regression model is a good fit for the data.

Next we will examine diagnostics of the logistic regression model.

### Figure 3

```{r, echo=FALSE,fig.width=5, fig.height=3,fig.align='center'}
cooks_distance <- cooks.distance(m4)
m4_dig=glm.diag(m4)
cd_plot=data.frame(obs=1:2698,cooksd=cooks_distance,leverage=m4_dig$h)
ggplot(data = cd_plot,aes(x=obs))+geom_line(aes(y=cd_plot$cooksd))+ylim(0,0.005)+ylab("Cook's Distance")+xlab("Observations")+labs(title = "Cook's Distance Plot")
```

The Cook's distance plot ranges from 0 to 0.004, showing that there are
few to no potential outliers. There are no clusters of high cooks'
distances, meaning that are no groups of observations that have strong
influence on the model. Additionally, we checked the variance inflation
factor (VIF) values of the predictors. They are all close to one,
indicating that there is low multicollinearity in the regression
analysis.

```{r, include=FALSE}
vif(m4)
```

## Results:

Based on our linear regression analysis, significant associations with
BMI percentile include birth weight, whether mother smoked during
pregnancy, seven days of physical activity, being Mexican American,
being Other Hispanic, being Asian, and being female with very low food
security. For every 1 pound increase in birth weight, BMI percentile
increases about 3 percentiles adjusting for all other variables.
Children with 7 days of physical activity have BMI percentiles about 14
percentiles lower compared to children with 0 days of physical activity
adjusting for all other variables. Additionally, the effect of very low
food security on BMI percentile is different for females and males.
Asian children have BMI percentiles about 11 percentiles lower compared
to White children adjusting for all other variables.

Based on our logistic regression analysis, the odds of being obese for
white Americans is about 1.49 times higher than Mexican Americans and
0.566 times the odds for Asians. The odds of obesity for children with
mothers who smoked during pregnancy are about 0.664 times the odds of
those without. The odds of obesity for children with physical activity 7
days a week are about 0.360 times the odds of those that do not have any
physically active days a week. The odds of obesity increase by about 19%
for a one pound increase in birth weight, and increase by about 5.8% for
a one meal increase in fast food per week.

### Figure 4

```{r,echo=FALSE,fig.height=3}
m2 <- lm(growth_percentile ~ sex+race+birthweight+smoke+food_sec+fastfood+activity1, data=data)

lm_model=summary(m2)
df=data.frame(lm_model$coefficients)
df$variable=rownames(df)
colnames(df)[2]="std_error"
df=df[df$variable!="(Intercept)",]
df$variable=c("Sex","Mexican American","Other Hispanic","Non-Hispanic Black","Non-Hispanic Asian","Other Race (Multi)","Birthweight","Smoked",
"Mariginal Security","Low Security","Very Low Food Sec","Fast Food","Activity 1-3","Activity 4-6","Activity 7")
df$type=ifelse(df$Estimate>0,"Positive","Negative")

log_model=summary(m4)
df1=data.frame(log_model$coefficients)
df1$variable=rownames(df1)
colnames(df1)[2]="std_error"
df1=df1[df1$variable!="(Intercept)",]
df1$variable=c("Sex","Mexican American","Other Hispanic","Non-Hispanic Black","Non-Hispanic Asian","Other Race (Multi)","Birthweight","Smoked",
"Mariginal Security","Low Security","Very Low Food Sec","Fast Food","Activity 1-3","Activity 4-6","Activity 7")
df1$type=ifelse(df1$Estimate>0,"Positive","Negative")

lm_plot=ggplot(df,aes(x=variable,y=Estimate))+geom_hline(yintercept = 0,linetype='dashed',color='red')+
  geom_point()+
  geom_errorbar(aes(ymin=Estimate-1.96*std_error,ymax=Estimate+1.96*std_error),width=0.2)+
  coord_flip()+xlab("Variables")+labs(subtitle = "Linear Regression Estimates") + theme(plot.subtitle = element_text(size = 9))

log_plot=ggplot(df1,aes(x=variable,y=Estimate))+geom_hline(yintercept = 0,linetype='dashed',color='red')+
  geom_point()+
  geom_errorbar(aes(ymin=Estimate-1.96*std_error,ymax=Estimate+1.96*std_error),width=0.2)+
  coord_flip()+xlab("Variables")+labs(subtitle = "Logistic Regression Estimates") + theme(plot.subtitle = element_text(size = 9))

ggarrange(lm_plot,log_plot,labels = c("A","B"))

```

## Conclusions:

Using linear regression, we identified factors associated with BMI
percentiles in children of ages 2-15. Additionally, using the results of
the linear regression analysis, we further modeled the probability of
obesity as a binary outcome using logistic regression. There are both
innate and environmental factors that are associated with BMI
percentiles and the log odds of obesity. As childhood obesity is a
serious problem that is often overlooked, it is important to further
study these factors and promote a healthier lifestyle among children In
the future, we would like to replicate our analysis after imputing data
for number of fast food meals/week using MICE (Multiple Imputation by
Chained Equation).

## References:

1.  Childhood Overweight & Obesity . Published 2023. Accessed November
    14, 2023.\
    [[https://www.cdc.gov/obesity/childhood/index.html\\\\](https://www.cdc.gov/obesity/childhood/index.html\){.uri}]([https://www.cdc.gov/obesity/childhood/index.html\\](https://www.cdc.gov/obesity/childhood/index.html)%7B.uri%7D){.uri}
2.  NHANES - About the National Health and Nutrition Examination Survey.
    Published 2023. Accessed November 14, 2023.\
    [[https://www.cdc.gov/nchs/nhanes/about_nhanes.htm\\\\](https://www.cdc.gov/nchs/nhanes/about_nhanes.htm\){.uri}]([https://www.cdc.gov/nchs/nhanes/about_nhanes.htm\\](https://www.cdc.gov/nchs/nhanes/about_nhanes.htm)%7B.uri%7D){.uri}
3.  Link to GitHub repository:
    <https://github.com/elenachun/625FinalProject>

## Contributions:

Elena Chun - Model fitting, diagnostics, part of results\
Aanya Mohan - Background, Data Overview, Conclusion\
Siwei Cao - Data cleaning, exploratory analysis
