---
title: "Analyzing the Factors Affecting BMI Percentiles of Children Aged 2-15"
authors: "Aanya Mohan, Elena Chun, Siwei Cao"
output: pdf_document
date: "2023-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:
Childhood obesity, also known as pediatric obesity, is a pressing issue that affects youth throughout the country. According to the Center for Disease Control (CDC), 
as of 2022, childhood obesity affects 1 in 5 children in the United States.1 There are a variety of factors that can lead to childhood obesity, from lifestyle to medication 
or genetic factors. Studies show that there may be a correlation between birth weight and the likelihood of childhood obesity. Despite sex differences, a higher birth weight 
increases the chances of developing obesity in adolescents. The National Health and Nutrition Examination Survey (NHANES) is “designed to address the health and nutritional 
status of adults and children in the United States”. This survey evaluates all aspects of a person’s health and nutritional habits. Factors recorded such as gender, age, food 
security, days physically active, mother smoked while pregnant, and number of meals from a fast food or pizza place, can be used to perform a regression analysis. This analysis 
may provide an explanation for childhood obesity trends, defined by birth weight. 

### Data Overview:
The data being used comes from The National Health and Nutrition Examination Survey (NHANES) conducted by the CDC for 2017-2020. The response variable for this analysis is 
growth percentile. The predictor of interest for this analysis is weight at birth, measured in pounds, reported for children ages 0-15 at the time of the survey. Other predictor 
variables include: gender, days physically active, mother smoked while pregnant, food security, race, and number of meals from a fast food or pizza place. These are all factors 
that are believed to explain trends in weight, and its effect on childhood obesity. Gender and race are demographic information reported by the survey participants. Days physically 
active is a metric measuring how many days of the week youths aged 2-17 are physically active for at least 60 minutes. Mother smoked while pregnant is a binary variable that 
assesses the smoking habits of the participants' biological mother. Food security is a categorical variable which rates the household food security on a scale of 1 to 4, where 1 is 
full food security and 4 is very low food security. Number of meals from a fast food or pizza place is the estimated number of meals not prepared at home, which are from a fast food 
restaurant or a pizza place, in a week. 

### Data Cleaning:
We intially wanted to use BMI as the response variable for our linear regression model. However, unlike adults, children are still growing. Therfore, what defines a 'healthy BMI' for 
children is highly dependent on their age. To combat this issue, we used the BMI information, as well as age, to calculate BMI percentiles based on equations provided by the CDC. 
Additionally, we combined birth weight in pounds (lbs) and birth weight in ounces (oz) to a single borthweight variable in pounds for easier interpretation. Something to note in this
section is the high number of missing data from the category "# of meals from fast food or pizza place". 
```{r, include=FALSE}
#Code to load data
```

### Exploratory Analysis:
Previous literature indicated that there would be a positive correlation between BMI percentile and birth weight. When plotting the entire data, it is unclear from this initial plot 
(Figure 1) whether there is a relationship or not. 
```{r, include=FALSE}
#Code for scatterplot of all data
```
After taking a random sample of 50 observations, we can see a slight positive correlation between birth weight and BMI percentile. 
```{r, include=FALSE}
#Code for random sample scatterplot
```

### Linear Regression Model:
The exporatory analysis of this data resulted in the following initial linear model:
$$Growth Percentile = \beta_{0} + \beta_{1}Sex + \beta_{2}Race + \beta_{3}BirthWeight + \beta_{4}Smoke + \beta_{5}FoodSecurity + \beta_{6}FastFood + \beta_{7}Activity + \varepsilon$$


### Diagnostics:

### Results:
Significant associations with BMI percentile include birth weight, whether mother smoked during pregnancy, seven days of physical activity, being Mexican American, being Other Hispanic,
being Asian, and being female with very low food security. For every 1 pound increase in birth weight, BMI percentile increases about 3 percentiles adjusting for all other variables.
Children with 7 days of physical activity have BMI percentiles about 14 percentiles lower compared to children with 0 days of physical activity adjusting for all other variables. 
Additionally, the effect of very low food security on BMI percentile is different for females and males. Asian children have BMI percentiles about 11 percentiles lower compared to 
White children adjusting for all other variables. 

### Conclusions:
Using linear regression, we identified factors associated with BMI percentiles in children of ages 2-15. There are both innate and environmental factors that are associated with 
BMI percentiles. As childhood obesity is a serious problem that is often overlooked, it is important to further study these factors and promote a healthier lifestyle among children
In the future, we would like to replicate our analysis after imputing data for number of fast food meals/week using MICE (Multiple Imputation by Chained Equation). 

